---
title: "Digit Classification Interactive Lesson"
output: learnr::tutorial
runtime: shiny_prerendered
tutorial:
  id: digit-classification-rsnns
  description: "Experiment with RSNNS to classify handwritten digits."
  version: 1.0.0
---

```{r setup, include=FALSE}

required_packages <- c("learnr", "RSNNS", "readr", "dplyr")

new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]

if(length(new_packages)) install.packages(new_packages)

library(learnr) #for lesson thing
library(RSNNS) #The Neural Net Bit
library(readr) #Import CSV
library(dplyr) 


train <- readr::read_csv("https://raw.githubusercontent.com/DefeatedWarrior/DataSci2/main/train.csv") #Pull the dataset from the Github

#This is where data gets cleaned

# Get total number of rows in the original training data
n <- nrow(train)

# Randomly sample 80% of row indices for training
train_idx <- sample(n, size = 0.8 * n)  # 80% for training, 20% for validation

# Split the original data into training and validation sets
train_data <- train[train_idx, ]        # Training subset (80%)
val_data <- train[-train_idx, ]         # Validation subset (remaining 20%)

# Prepare training features:
# - Drop the label column
# - Convert to matrix
# - Normalize pixel values to [0, 1]
x_train <- as.matrix(train_data[, -1]) / 255

# Prepare training labels as one-hot vectors for RSNNS
y_train <- decodeClassLabels(train_data$label)

# Prepare validation features and labels
x_val <- as.matrix(val_data[, -1]) / 255     # Same normalization
y_val <- val_data$label                      # Keep raw labels for accuracy check


```

```{r js_ui, echo=FALSE, results='asis'}
tags$script(HTML("

  // Register a custom message handler named 'initPixelClicks'
  // This is called from the server with: session$sendCustomMessage(...)
  Shiny.addCustomMessageHandler('initPixelClicks', function(message) {

    // Remove any previous click event listeners on pixel divs (avoid duplicates)
    $(document).off('click', \"div[id^='pixel_']\");

    // Add new click event listener to any divs with IDs starting with 'pixel_'
    $(document).on('click', \"div[id^='pixel_']\", function() {
      
      // Send the clicked pixel's ID back to the Shiny input system
      Shiny.setInputValue('clicked_pixel', this.id, {priority: 'event'});
    });
  });

"))
```

## Welcome

Welcome to the interactive tutorial on digit classification using a neural network in R.

---

## Choose Model Architecture

Use the slider below to adjust the number of neurons in the hidden layer. (64 or less for this lesson.) Press train now button and proceed to next section.

```{r input_panel}
sliderInput("hidden_size", "Number of hidden neurons:", min = 4, max = 256, value = 64, step = 4)

actionButton("train_button", "Train Model")

```

---

## Train the Neural Network

Wait here until Training completed successfully!

```{r train_model, context="server"}


# Define a reactive expression for training the model
# This re-trains the model each time the "Train Model" button is clicked
model <- eventReactive(input$train_button, {
  
  set.seed(123)  # Ensures reproducibility of training results

  # Train a Multi-Layer Perceptron (MLP) using RSNNS
  # Arguments:
  # - x_train: input features (matrix, scaled to [0, 1])
  # - y_train: one-hot encoded labels
  # - size: number of hidden neurons (from user-selected slider)
  # - learnFunc: standard backpropagation algorithm
  # - learnFuncParams: learning rate of 0.1
  # - maxit: run for 50 iterations (epochs)
  mlp(
    x_train,
    y_train,
    size = input$hidden_size,
    learnFunc = "Std_Backpropagation",
    learnFuncParams = c(0.1),
    maxit = 50
  )
})

# Output message to indicate whether training completed successfully
output$train_status <- renderText({
  req(input$train_button)  # Only proceed if the train button has been clicked

  isolate({  # Prevent reactivity from triggering unnecessary re-runs
    if (!is.null(model())) {
      "Training completed successfully!"  # Model is trained and ready
    } else {
      "Model training failed."  # Something went wrong
    }
  })
})



```

```{r train_status_ui}
textOutput("train_status") 

#Actually outputing that (Could add this to the other page but thats a major change and I don't want to tempt the vengeful programming gods)
```

---

## Test Model

```{r inpred}
# Heading for the digit drawing section
h3("Draw a Digit")

# Inject custom CSS styles to remove spacing between grid pixels
tags$style(HTML("
  /* Ensure each individual pixel (div) has no margin/padding */
  #draw_grid_ui div[id^='pixel_'] {
    box-sizing: border-box;
    margin: 0;
    padding: 0;
  }

  /* Also remove padding/margin from rows in the grid container */
  #draw_grid_ui > div {
    margin: 0;
    padding: 0;
  }
"))

# Display the 28x28 interactive drawing grid
uiOutput("draw_grid_ui")

# Button to clear the drawing (reset all pixels to white)
actionButton("clear_grid", "Clear Drawing")

# Button to run the model on the current drawing
actionButton("predict_drawn", "Predict Drawn Digit")

# Output area to display predicted digit (e.g., "Predicted digit is: 7")
verbatimTextOutput("draw_prediction")


```

```{r prediction, context="server"}
# Create a reactive 28x28 matrix initialized to 0s.
# This will represent the drawn digit (each cell is a pixel: 0 = off, 1 = on).
drawing <- reactiveVal(matrix(0, nrow = 28, ncol = 28))

# Render the 28x28 drawing grid UI using a nested set of <div> elements.
output$draw_grid_ui <- renderUI({
  mat <- drawing()  # Get the current drawing matrix
  
  # Create 28 rows
  rows <- lapply(1:28, function(row) {
    
    # For each row, create 28 square "pixel" cells
    cells <- lapply(1:28, function(col) {
      pixel_id <- paste0("pixel_", row, "_", col)  # Unique ID for each pixel
      
      # Create a small square div with black or white background depending on state
      div(
        id = pixel_id,
        style = paste0(
          "width: 15px; height: 15px; border: 1px solid #ddd; cursor: pointer; background-color:",
          ifelse(mat[row, col] == 1, "black", "white"),  # Black if ON, white if OFF
          "; margin: 0; padding: 0;"
        )
      )
    })
    
    # Wrap each row of cells in a horizontal flex container
    div(style = "display: flex; margin: 0; padding: 0;", cells)
  })
  
  # Send a message to JS side to initialize pixel click event listeners
  session$sendCustomMessage("initPixelClicks", list())
  
  # Return the full grid of rows
  tagList(rows)
})

# React to a pixel being clicked (JS sends back the clicked pixel's ID)
observeEvent(input$clicked_pixel, {
  coords <- strsplit(input$clicked_pixel, "_")[[1]][-1]  # Extract row/col from ID
  row <- as.integer(coords[1])
  col <- as.integer(coords[2])
  
  mat <- drawing()                   # Get current drawing
  mat[row, col] <- 1 - mat[row, col] # Toggle between 0 and 1 (flip pixel state)
  drawing(mat)                       # Update the reactive matrix
})

# Clear the drawing (set all pixels to 0)
observeEvent(input$clear_grid, {
  drawing(matrix(0, 28, 28))
})

# When "Predict" is clicked, run the model on the current drawing
observeEvent(input$predict_drawn, {
  req(model())                          # Ensure model is trained
  
  input_vec <- as.numeric(t(drawing())) # Flatten the matrix to a vector row-wise
  pred_probs <- predict(model(), matrix(input_vec, nrow = 1)) # Get prediction
  pred_digit <- which.max(pred_probs) - 1 # Convert one-hot to digit (0-indexed)
  
  # Display the predicted digit
  output$draw_prediction <- renderText({
    paste("Predicted digit is:", pred_digit)
  })
})


```

```{r predict_ui}
verbatimTextOutput("prediction")
```

---

## Well Done!

You've built and trained a neural network to classify digits using RSNNS in R. Try adjusting the architecture to see how performance changes! Once done, Continue to the next section to access the quiz

---

## Quiz: What Did You Learn?

Please complete the quiz on Microsoft Forms https://forms.office.com/r/eL9rAr5GQm
